The missing puzzle piece: contextual insights for enhanced pharmaceutical supply chain forecasting

### 2025-08-04 manuscript/01-introduction.qmd
3RD REVISION

Demand forecasting in pharmaceutical supply chains is a highly complex process, shaped by several factors. These include the quality of available data, the specific position within the supply chain, and external influences such as market trends, regulatory changes, and disruptive events such as pandemics, conflicts, flooding [@schneider2010pharmaceutical; @hyndman2018forecasting]. Recently, major donors  such as Bill & Melinda Gates Foundation have invested in data collection technologies in sub-Saharan Africa, recognizing the potential of data to provide valuable insights. This investment is essential and offers significant benefits for improving supply chain delivery. However, much of the data being collected focuses on transactional data, such as consumption and sale. Almost all platforms collecting data tend to overlook the critical contextual factors and events that influence these transactions—factors that are just as crucial for accurate forecasting. This presents a major barrier, as understanding the broader context, including administration procedures, local policies, disruptions, conflicts, and so on is vital for building reliable forecasting models. This gap—known as business context and domain knowledge—requires urgent attention in Pharmaceutical supply chain forecasting. Without incorporating this knowledge, any forecasting model is unlikely to produce reliable results. We must prioritize collecting and integrating this contextual information to improve the accuracy of demand forecasts in pharmaceutical supply chains. Combining statistical models with domain insights offers a more refined forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview]. Domain knowledge enables analysts to incorporate relevant events, such as supply chain disruptions, into the forecasting model. This integration is particularly beneficial in fields like healthcare, where expert insights can align algorithms with real-world applications [@dash2022review].

@@

This research uses consumption data from 33 health commodities, covering the period from December 2017 to July 2022, provided by the Ethiopian Pharmaceutical Supply Service (EPSS), to develop models for six-month-ahead forecasting. Initially, we implement univariate models that rely solely on consumption data. Following this, we conduct exploratory data analysis and engage in interviews with EPSS experts to gather domain knowledge on events influencing consumption patterns. Through these interviews, the experts identified three key factors collectively impacting consumption: stock replenishment period, fiscal calendar, and the seasonal malaria. We then develop forecasting models that incorporate the data gathered through the interview process and compare their performance with univariate models that rely solely on consumption data.

>>>>

Demand forecasting in pharmaceutical supply chains is a highly complex process, shaped by several factors. These include the quality of available data, the specific position within the supply chain, and external influences such as market trends, regulatory changes, and disruptive events such as pandemics, conflicts, flooding [@schneider2010pharmaceutical; @hyndman2018forecasting]. Recently, major donors  such as Bill & Melinda Gates Foundation have invested in data collection technologies in sub-Saharan Africa, recognizing the potential of data to provide valuable insights. This investment is essential and offers significant benefits for improving supply chain delivery. However, much of the data being collected focuses on transactional data, such as consumption, distribution, and sale. Almost all platforms collecting data tend to overlook the critical contextual factors and events that influence these transactions—factors that are just as crucial for accurate forecasting. This presents a major barrier, as understanding the broader context, including administration procedures, local policies, disruptions, conflicts, and so on is vital for building reliable forecasting models. This gap — known as business context and domain knowledge — requires urgent attention in pharmaceutical supply chain forecasting. Without incorporating this knowledge, any forecasting model is unlikely to produce reliable results. We must prioritize collecting and integrating this contextual information to improve the accuracy of demand forecasts in pharmaceutical supply chains. Integrating statistical models with domain knowledge leads to a more accurate and context-aware forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview]. Domain knowledge enables analysts to incorporate relevant events, such as supply chain disruptions, into the forecasting model. This integration is particularly beneficial in fields like healthcare, where expert insights can align algorithms with real-world applications [@dash2022review].

@@

This research uses distribution data from 33 health commodities, covering the period from December 2017 to July 2022, provided by the Ethiopian Pharmaceutical Supply Service (EPSS), to develop models for six-month-ahead forecasting. Initially, we implement univariate models that rely solely on distribution data. Following this, we conduct exploratory data analysis and engage in interviews with EPSS experts to gather domain knowledge on events influencing distribution patterns. Through these interviews, the experts identified three key factors collectively impacting distribution: stock replenishment period, fiscal calendar, and the seasonal malaria. We then develop forecasting models that incorporate the data gathered through the interview process and compare their performance with univariate models that rely solely on distribution data.

### 2025-08-04 manuscript/02-literature.qmd
3RD REVISION

Accurate sales forecasting is crucial in the pharmaceutical industry, where it directly impacts profit maximization, cost minimization, and the ability to respond to market changes. Forecasting in healthcare not only influences clinical decisions but also plays a pivotal role in supply chain management, ensuring that the right drugs are available when needed. The challenge lies in balancing customer demand with inventory costs, which is particularly complex in pharmaceuticals due to factors like short shelf life and quality constraints [@gupta2000mid; @makridakis2020forecasting].

In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies using a combination of simple and advanced methods to enhance accuracy [@nikolopoulos2016forecasting;@zhu2021demand;@anusha2014demand]. Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods like the Holt-Winters method, Single Moving Average (SMA), and Single Exponential Smoothing (SES), which are simpler but effective for certain applications. In contrast, other researchers explored more advanced approaches, including hybrid models and machine learning techniques, to address complex forecasting challenges, as seen in the studies by @siddiqui2022hybrid, @de2021lead, @kim2015demand, @candan2014demand, and @ribeiro2017improving.

Early research in pharmaceutical forecasting primarily focused on traditional statistical methods. These methods, including ARIMA, exponential smoothing, and moving averages, were widely used due to their simplicity and effectiveness in relatively stable environments [@zahra2019forecasting]. These models typically assess forecast accuracy using metrics like Mean Absolute Percentage Error (MAPE), Mean Squared Error (MSE), and Root Mean Squared Error (RMSE). For example, @newberne2006holt applied the Holt-Winters method to forecast healthcare data, focusing on prescription trends. Their findings highlighted the method's utility in short-term planning and resource management, validated through metrics like MAPE, MAD, and MSD. Similarly, @restyana2021analysis compared Single Moving Average (SMA) and Single Exponential Smoothing (SES) in drug demand forecasting, concluding that SES provided more accurate results, as indicated by lower MAD and MSE values.

Recognizing the limitations of traditional models, particularly in capturing the complexities of pharmaceutical demand (e.g., seasonality, external influences), researchers began developing hybrid models. These models combine linear methods like ARIMA with nonlinear techniques such as neural networks, aiming to improve forecasting accuracy by addressing diverse data patterns. @khalil2014intelligent introduced a hybrid model that integrated ARIMA with neural networks. Their study demonstrated that this approach significantly outperformed traditional models, especially in scenarios with limited historical data. The hybrid model's effectiveness was confirmed by improvements in key performance indicators like MAPE and RMSE.

The pharmaceutical industry's increasing complexity and the availability of large datasets have led to the adoption of advanced machine learning techniques. These methods, such as Support Vector Regression (SVR), Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer greater flexibility and predictive power. @van2021using showed that incorporating downstream information into machine learning models, like LASSO and SVR, greatly enhances forecast accuracy. Their study on multi-echelon supply chains revealed that advanced models, particularly those integrating external variables, consistently produced lower forecast errors, as measured by AvgRelRMSE. Similarly, @rathipriya2023demand compared shallow neural networks and deep learning models for drug demand forecasting. The study found that shallow models outperformed deep learning models for most drug categories, while ARIMA was more effective for the remaining categories. This suggests that no single model is universally optimal, and model selection should be context-specific.

@@

1. Despite significant investment in data collection technologies and logistics management systems, these systems primarily collect transactional data, such as consumption. However, no data is collected on events that influence consumption variability—information that would be crucial for building reliable forecasting models. No studies to date have explored how domain expertise could enhance pharmaceutical demand forecasting.

2. Current research on forecasting for pharmaceutical supply chain predominantly focuses on generating point forecasts. There is a lack of studies that consider the entire forecast distribution of monthly consumption, which would better capture the uncertainty of future demand and provide a valuable risk management tool for decision-makers.

>>>>

Accurate demand forecasting is crucial in the pharmaceutical industry, where it directly impacts profit maximization, cost minimization, and the ability to respond to market changes. Forecasting in healthcare not only influences clinical decisions but also plays a pivotal role in supply chain management, ensuring that the right drugs are available when needed. The challenge lies in balancing customer demand with inventory costs, which is particularly complex in pharmaceuticals due to factors like short shelf life and quality constraints [@gupta2000mid; @makridakis2020forecasting].

A range of methods is used in pharmaceutical supply chain forecasting, from traditional statistical models to advanced machine learning techniques, with many studies combining both simple and advanced approaches to improve accuracy [@nikolopoulos2016forecasting;@zhu2021demand;@anusha2014demand]. Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods. In contrast, other researchers explored more advanced approaches, including hybrid models and machine learning techniques, to address complex forecasting challenges, as seen in the studies by @siddiqui2022hybrid, @de2021lead, @kim2015demand, @candan2014demand, and @ribeiro2017improving.

Early research in pharmaceutical forecasting primarily focused on classical statistical methods. These methods, including ARIMA (AutoRegressive Integrated Moving Average), exponential smoothing, and moving averages, were widely used due to their simplicity and effectiveness in relatively stable environments [@zahra2019forecasting]. These models typically assess forecast accuracy using metrics like Mean Absolute Percentage Error (MAPE), Mean Squared Error (MSE), and Root Mean Squared Error (RMSE). For example, @newberne2006holt applied the Holt-Winters method to forecast healthcare data, focusing on prescription trends. Their findings highlighted the method's utility in short-term planning and resource management, validated through metrics like MAPE, MAD, and MSD. Similarly, @restyana2021analysis compared  Simple Moving Average (SMA), and Single Exponential Smoothing (SES) in drug demand forecasting, concluding that SES provided more accurate results, as indicated by lower MAD and MSE values.

Recognizing the limitations of traditional models—particularly their inability to capture the complexities of pharmaceutical demand, such as seasonality and external influences—researchers have increasingly turned to more advanced modeling approaches [@khalil2014intelligent]. Furthermore, the growing complexity of the pharmaceutical industry, coupled with the availability of large datasets, has accelerated the adoption of advanced machine learning techniques. These methods, such as Support Vector Regression (SVR), Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer greater flexibility and predictive power. @van2021using showed that incorporating downstream information into machine learning models, like LASSO and SVR, greatly enhances forecast accuracy. Their study on multi-echelon supply chains revealed that advanced models, particularly those integrating external variables, consistently produced lower forecast errors, as measured by AvgRelRMSE. Similarly, @rathipriya2023demand compared shallow neural networks and deep learning models for drug demand forecasting. The study found that shallow models outperformed deep learning models for most drug categories, while ARIMA was more effective for the remaining categories. This suggests that no single model is universally optimal, and model selection should be context-specific.

@@

1. Despite significant investment in data collection technologies and logistics management systems, these systems primarily collect transactional data, such as consumption and distribution. However, no data is collected on events that influence their variability — information that would be crucial for building reliable forecasting models. No studies to date have explored how domain expertise could enhance pharmaceutical demand forecasting.

2. Current research on forecasting for pharmaceutical supply chain predominantly focuses on generating point forecasts. There is a lack of studies that consider the entire forecast distribution, which would better capture the uncertainty of future demand and provide a valuable risk management tool for decision-makers.
